---
title: "Statistical test for NO difference"
subtitle: "Equivalence Testing for Psychological Research: A Tutorial"
citation: "Advances in Methods and Practices in Psychological Science, 2018"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: powerpoint_presentation
---

- Statistical (hypothesis) tests are oriented to test for differences
- The null hypothesis states "no difference"; *rejecting* the null hypothesis implies there is a difference.
- p-value is the probability of seeing the observed magnitude change (or higher) if the null were true.
- How do we do the inverse? How do we get a p-value for two things being equal?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(readxl)
library(reshape2)
library(ggplot2)
library(TOSTER)

```

```{r confint, echo=F}

confint = function(data, conf.int=.95)
{
  n = length(data)
  m = mean(data)
  s = sd(data)
  se = s/sqrt(n)
  ci = se * qt(conf.int/2 + .5, df=n-1)
  return( c(-ci,0,ci) + m)
}

pooled_sd = function(x1,x2) {
  sd1 = sd(x1)
  sd2 = sd(x2)
  n1 = length(x1)
  n2 = length(x2)
  
  sqrt( ((n1-1)*sd1^2 + (n2-1)*sd2^2) / (n1+n2-2)  )
}

welch_s_delta = function(x1,x2) {
  sd1 = sd(x1)
  sd2 = sd(x2)
  n1 = length(x1)
  n2 = length(x2)
  sqrt( (sd1^2/n1) + (sd2^2/n2))
}

```

## Are these two samples the same? 

```{r data, echo = FALSE,fig.width=3.35}
xls = read_xlsx('21820 NS34A IT PRVABC59.xlsx',2)
# use a log2 of the raw data, so that later we can express a 2 fold change as
# one integral unit on the X-axis
BF2Log2 = log(xls$BF2,2)
BF4Log2 = log(xls$BF4,2)

data = data.frame(control=BF2Log2, antiNS3_4A=BF4Log2)
tidydata = melt(
  data,
  measure.vars = c(1, 2),
  variable.name = "group",
  value.name = "viral titre"
)
baseplot <- ggplot(tidydata, aes(x = group, y = `viral titre`)) + 
  scale_y_continuous(limits = c(15, 24)) + 
  labs(y="Log2 (Plaque Forming Units/mL) ZIKV", x="experimental group") +     
    geom_point(aes(fill=group), color='black', stroke=1, size=2, shape=21,
    position = position_jitter(seed=0,width=.15)) +    
    scale_fill_manual(values=c("#F3B77C","#244AFC")) +
  theme_classic() + theme(legend.position="none") +  labs(title="Viral titre in infected mosquitos")


plot1 <- baseplot +
  geom_boxplot(outlier.shape = NA, fill="white",alpha=.2,width=.5) +
  labs(subtitle="line - median, box - interquartile range")
      
  
print(plot1)
```


### Same plot, mean and confidence intervals
``` {r ci, echo=FALSE, fig.width=3.35}
plot2 <- baseplot +
  stat_summary(fun.data = "mean_se", fill="white", alpha=.2, size=.5, color = "red", geom = "crossbar",width=.5) +
  labs(subtitle="Mean, 95% confidence interval")
print(plot2 + ylab(element_blank()))
```

### T-test (unequal means)
``` {r t-test}
t_unequal_means = t.test(BF2Log2,BF4Log2)
diff_in_means = abs( diff( c(mean(BF2Log2), mean(BF4Log2) ) ) )
sdw = welch_s_delta(BF2Log2,BF4Log2)
cv = qt(.975, t_unequal_means$parameter)*sdw # critical value for two-tailed, alpha=.05
plot_t = function(x,mu,df) { dt((x-mu)/sdw, df=df)}
xleft=-1;xright=1
xseries=seq(xleft,xright,length.out=250)
yseries=plot_t(xseries,mu=0,df=t_unequal_means$parameter)
gg=data.frame(x=xseries,y=yseries,  
              yleft=plot_t(xseries,  
                    mu=-1,
                    df=t_unequal_means$parameter),
               yright=plot_t(xseries,  
                    mu=1,
                    df=t_unequal_means$parameter)
               )
ggplot(gg) + theme_classic() +
  coord_cartesian(ylim=c(0,.4), xlim=c(-1.1,1.1), expand=F) + 
  geom_hline(yintercept=0) +
  geom_vline(xintercept=c(-1,1),linetype=2) +
  geom_vline(xintercept=c(0),linetype=2,color="grey") +
  geom_vline(xintercept=diff_in_means, color='red') + 
  xlab("difference in means") +
  ylab("probability distribution (t statistic)") -> distbaseplot
  
  distbaseplot + stat_function(fun=plot_t,args=list(mu=0,df=t_unequal_means$parameter)) +
  geom_linerange(data=gg[abs(gg$x)>cv,], aes(x,ymin=0,ymax=y), color='blue') +
  labs(title="t-test for unequal means", subtitle=sprintf("Observed difference: %.3f (red line) ≠ 0",diff_in_means ),caption=sprintf("Critical values (dashed blue line): |%.3f|; p-value=%.3f", cv, t_unequal_means$p.value))
```

### If a p-value < .05 means there is a difference, doesn't a high p-value(say, > .95) mean no difference?

- Not necessarily. P-values are oriented to reject a null hypothesis, not accept one.
- How do we define (and possibly reject) a null hypothesis of **change**? The classical tests are null hypothesis = no change.
- How do we test for ***no difference?*** 

### Testing for equivalence 

- Step 1: Pre-define the smallest effect of interest. We will discuss the choice of smallest effect size later. 
- This dataset: 2-fold change is the benchmark in the lab. In log2 space, this is ±1.
- Step 2: Define the lower and upper equivalence bounds based on Step 1.
- Step 3: Perform a one-sided test against each boundary: 1) observed > -1; 2) observed < 1;
- Step 4: If both tests pass (p1 < alpha *and* p2 < alpha), *conclude **no difference** between the groups.* Report the higher p-value.

``` {r visualize-cis}
cv2 = qt(.05, t_unequal_means$parameter)*sdw + 1 # a critical value for mu=1
between = function(x, xmin, xmax) {
  (x > xmin) & (x < xmax)
}

distbaseplot + 
  #geom_vline(xintercept=c(-1)*cv2, color='#8080FF',linetype=2,size=.4) + 
  #geom_vline(xintercept=c(1)*cv2, color='blue',linetype=2,size=.4) + 
  stat_function(fun=plot_t,args=list(mu=-1,df=t_unequal_means$parameter),xlim=c(-1.2,1.2)) +
  stat_function(fun=plot_t,args=list(mu=1,df=t_unequal_means$parameter),xlim=c(-1.2,1.2)) +
  geom_linerange(data=gg[between(gg$x,-cv2,0),], aes(x,ymin=0,ymax=yleft), color='#8080FF') +
  geom_linerange(data=gg[between(gg$x,0,cv2),], aes(x,ymin=0,ymax=yright), color='blue') +
  labs(title="Two one-sided t-tests for equivalence", subtitle=sprintf("[ -1 < %.3f < 1 ]",diff_in_means ),caption=sprintf("Red line is inside the rejection regions of both tests")) +
  xlab("difference in means") +
  ylab("probability distribution (t statistic)")

```

## TOST: Two One-sided Tests

R-Package: TOSTER. Plot is simpler.

```{r Two One-Sided Test, echo = TRUE}

TOSTtwo.raw( mean(BF2Log2), mean(BF4Log2), 
         sd(BF2Log2), sd(BF4Log2), 
         length(BF2Log2), length(BF4Log2),  
         low_eqbound = -1, 
         high_eqbound = 1, alpha=.05) 


```

## Choice of smallest effect size of interest (SESOI)

Tutorial (Lakëns et al., 2018) discussion.

### Just-noticeable difference. Burriss et al., 2015:

Research question: Do ovulating women signal men by becoming more flush in the face?

- NIH funded research?
- Maybelline?

Just-noticeable difference: Change in face redness perceptable by men.

Authors' conclusions: women do become more flush, but not enough to be discernable by the naked eye. (Maybelline withdraws funding)

### Statistical Power, no laughing matter

  - What **effect size** can we detect at alpha=.05, sample size n, power=.8?
  - **Effect size** in units of *Cohen's d.*
  - Run TOST using **effect size**, rather than raw value.
